Along-Track Acceleration of Tropical Cyclones: A Proxy For Extratropical Interactions

### 2020-10-03 paper/manuscript.tex

 As our intent here was to examine the motion of tropical storms, in subsequent sections unless explicitly stated,  we focused on the track data prior to the completion of ET. To ascertain whether a tropical storm underwent ET, we used the \emph{nature} designation in the IBTraCS database that relies on the judgement of the forecasters from one or more agencies responsible for the ocean basin.  Admittedly, there will be some subjectivity in this designation. An alternative would be to employ a metric such as the cyclone phase space \citep{Hart2003}, usually calculated using meteorological fields from global numerical weather prediction models (e.g., reanalysis data). As we wish to be independent of modeled products to characterize the storm, we choose to use the forecaster designated storm nature. Furthermore, \cite{BCSEH2019a} found that the phase space calculation was sensitive to the choice of the reanalysis model. Occasionally, the nature of the storm is not recorded (NR) or, if there is an inconsistency among the agencies, it is designated as mixed (MX). In the north Atlantic, only a small fraction of track data ($<0.6\%$) in the IBTrACs is designated as NR or MX. However, the fraction is higher in the North Pacific ($\approx 15\%$), and we return to this point in a subsequent section where it could introduce some uncertainty in our conclusion.

@@

During 1966--2019, a total of 813 storms in the North Atlantic and 1782 storms in the western North Pacific passed the lifetime threshold of 2 days. Figures \ref{fig:distNA} and \ref{fig:distWP} shows some basic statistics for TC speed and accelerations as a function of latitude. TC locations were sorted in $10^\circ$-wide latitude bins for this. Tables \ref{tab:AC1}-\ref{tab:AT2} provide the same data. The average speed of all North Atlantic storms (including over-land) is about 21 km/hr. The same for the western Pacific storms is about 19 km/hr. As expected, TC speed clearly varies with latitude. It is lower in the subtropics as compared to other latitudes. It increases sharply in the vicinity of $40^{o}$ N.

@@

\begin{table*}[t]
  \caption{Trends in Speed (km hr$^{-1}$ year$^{-1}$) of Atlantic. All storms refers to all instances of a
    system recorded in the IBTraCs. ET refers to storm ntaure designated as extratropical, while NR refers to instances
    when the storm nature was not recorded. }

@@

\begin{table*}[t]
\caption{Trends in Speed (km hr$^{-1}$ year$^{-1}$), tangential and curvature acceleration  (km hr$^{-1}$ day$^{-1}$ year$^{-1}$) of all Atlantic non-ET storms as a function of latitude: N refers to number of 3-hourly track positions in each latitude-bin over the period 1966--2019. Storm positions corresponding to 
nature labels "ET" and "NR" were excluded.}

\begin{tabular}{cc|ccc|ccc|ccc}
\tophline
& & \multicolumn{3}{|c|}{Speed} & \multicolumn{3}{c|}{Tang. Accel} & \multicolumn{3}{c}{Curv. Accel}\\
\middlehline
Latitude & N & Mean & Median & Std Dev. & Mean & Median &  Std Dev.& Mean & Median & Std Dev.\\

\bottomhline
\end{tabular}
%\belowtable{} % Table Footnotes
\label{tab:ATLSA}
\end{table*}

@@

\begin{figure*}[t]
\includegraphics[width=7cm]{altl_speed_trend.png}
\caption{Annual-mean speed of Atlantic storms, along with the linear trend. The purple curve is for all storms while the green curve excludes instances when
the storm was classified as ET or NR.}\label{fig:atl_speed}
\end{figure*}
%========================================================

>>>>

 As our intent here was to examine the motion of tropical storms, in subsequent sections unless explicitly stated,  we focused on the track data prior to the completion of ET. To ascertain whether a tropical storm underwent ET, we used the \emph{nature} designation in the IBTraCS database that relies on the judgement of the forecasters from one or more agencies responsible for the ocean basin.  Admittedly, there will be some subjectivity in this designation. An alternative would be to employ a metric such as the cyclone phase space \citep{Hart2003}, usually calculated using meteorological fields from global numerical weather prediction models (e.g., reanalysis data). As we wish to be independent of modeled products to characterize the storm, we choose to use the forecaster designated storm nature. Furthermore, \cite{BCSEH2019a} found that the phase space calculation was sensitive to the choice of the reanalysis model. Occasionally, the nature of the storm is not recorded (NR) or, if there is an inconsistency among the agencies, it is designated as mixed (MX). In the north Atlantic, only a small fraction of track data ($\approx 0.5\%$) in the IBTrACs is designated as NR or MX. However, the fraction is higher in the North Pacific ($\approx 14\%$), and we return to this point in a subsequent section where it could introduce some uncertainty in our conclusion for this basin.

@@

During 1966--2019, a total of 688 storms in the North Atlantic and 1676 storms in the western North Pacific passed the lifetime threshold of 2 days.

Figures \ref{fig:distNA} and \ref{fig:distWP} shows some basic statistics for TC speed and accelerations as a function of latitude. TC locations were sorted in $10^\circ$-wide latitude bins for this. Tables \ref{tab:AC1}-\ref{tab:AT2} provide the same data. The average speed of all North Atlantic storms (including over-land) is about 21 km/hr. The same for the western Pacific storms is about 19 km/hr. As expected, TC speed clearly varies with latitude. It is lower in the subtropics as compared to other latitudes. It increases sharply in the vicinity of $40^{o}$ N.

@@

\begin{table*}[t]
  \caption{Trends in Speed (km hr$^{-1}$ year$^{-1}$). \emph{All storms} refers to all instances of a
    system recorded in the IBTraCs. ET refers to storm ntaure designated as extratropical, while NR refers to instances
    when the storm nature was not recorded. }

@@

\begin{table*}[t]
  \caption{Trends in rapid tangential acceleration (km hr$^{-1}$ day$^{-1}$ year$^{-1}$) of Atlantic tropical cyclones. Storm instances  classified as ET or NR are excluded. Three cut-offs for defining rapid are used: Values exceeeding 90th, 85th and 60th percentile of all acceleration. }

\begin{tabular}{|c|cccc|cccc|cccc|}
\hline
%& \multicolumn{4}{|c|}{1966--2019}  & \multicolumn{4}{|c|}{1966--2019}\\
& \multicolumn{4}{|c|}{ 90th percentile } & \multicolumn{4}{|c|}{80th percentile }& \multicolumn{4}{|c|}{60th percentile}\\
Atlantic  & \multicolumn{2}{c}{LR} & \multicolumn{2}{c|}{MK-TS}& \multicolumn{2}{c}{LR} & \multicolumn{2}{c|}{MK-TS} & \multicolumn{2}{c}{LR} &  \multicolumn{2}{c|}{MK-TS}\\
(Excluding ET,NR) & Trend & P-value & Trend & P-value & Trend & P-value & Trend & P-value & Trend & P-value & Trend & P-value\\
\hline
Full basin  & -0.097 & 0.01 & -0.091 & 0.01 & -0.088 &  <.01 & -0.084 &  <.01 & -0.051 & 0.01 & -0.050 & 0.02 \\
	  5--25 & -0.094 & 0.00 & -0.102 & 0.00   & -0.081 & 0.00 & -0.080 & 0.00 & -0.046 & 0.01 & -0.047 & 0.02\\
	 10--30 & -0.080 & 0.01 & -0.082 & 0.00   & -0.079 & 0.00 & -0.072 & 0.00 & -0.048 & 0.01 & -0.049 & 0.01\\
	 15--35 & -0.086 & 0.02 & -0.079 & 0.01   & -0.084 & 0.00 & -0.079 & 0.00 & -0.043 & 0.03 & -0.045 & 0.02\\
	 20--40 & -0.113 & 0.01 & -0.110 & 0.01   & -0.088 & 0.01 & -0.086 & 0.02 & -0.046 & 0.06 & -0.051 & 0.06\\
	 25--45 & -0.099 & 0.08 & -0.090 & 0.05   & -0.075 & 0.10 & -0.065 & 0.11 & -0.061 & 0.07 & -0.047 & 0.10\\
	 30--50 & -0.124 & 0.06 & -0.119 & 0.04   & -0.097 & 0.08 & -0.099 & 0.11 & -0.078 & 0.08 & -0.061 & 0.18\\
\hline
\hline
\end{tabular}
%\belowtable{} % Table Footnotes
\label{tab:ATLAC}
\end{table*}

\begin{table*}[t]
  \caption{Trends in rapid tangential acceleration (km hr$^{-1}$ day$^{-1}$ year$^{-1}$) of Atlantic tropical cyclones. All storm instance, including classified as NR and ET are considered here. Three cut-offs for defining rapid are used: Values exceeeding 90th, 85th and 60th percentile of all acceleration. }

\begin{tabular}{|c|cccc|cccc|cccc|}
\hline
%& \multicolumn{4}{|c|}{1966--2019}  & \multicolumn{4}{|c|}{1966--2019}\\
& \multicolumn{4}{|c|}{ 90th percentile } & \multicolumn{4}{|c|}{80th percentile }& \multicolumn{4}{|c|}{60th percentile}\\
Atlantic  & \multicolumn{2}{c}{LR} & \multicolumn{2}{c|}{MK-TS}& \multicolumn{2}{c}{LR} & \multicolumn{2}{c|}{MK-TS} & \multicolumn{2}{c}{LR} &  \multicolumn{2}{c|}{MK-TS}\\
(Excluding ET,NR) & Trend & P-value & Trend & P-value & Trend & P-value & Trend & P-value & Trend & P-value & Trend & P-value\\
\hline
Full basin &  -0.075 & 0.06 & -0.080 & 0.05 & -0.055 & 0.05 & -0.055 & 0.06  & -0.026 & 0.22 & -0.024 & 0.32\\

\hline
\hline
\end{tabular}
%\belowtable{} % Table Footnotes
\label{tab:ATLAC}
\end{table*}

@@

%========================================================
%Figure 
%========================================================

\begin{figure*}[t]
  \includegraphics[width=7cm]{atl_speed_trend.png}
  \includegraphics[width=7cm]{wpa_speed_trend.png}

\caption{Annual-mean speed of storms in the (a) North Atlantic; and (b) western North Pacific, along with the linear trend. The purple curve is for all storms while the green curve excludes instances when
the storm was classified as ET or NR.}\label{fig:atl_speed}
\end{figure*}
%========================================================

%========================================================
%Figure 
%========================================================

\begin{figure*}[t]
  \includegraphics[width=7cm]{atl_tang_no_etnr.png}
   \includegraphics[width=7cm]{atl_tang_20_40_no_etnr.png}

  \caption{Annual-mean tangential acceleration exceeding thresholds of 60th, 80th and 90th percentile values for (a) Entire Atlantic; and (b) Latitude range 20-40$^o$N. Instances of storms classified as NR and ET were excluded. The linear trend for each timeseries is shown by the straight line.}\label{fig:atl_speed}
\end{figure*}
%========================================================
